Data Migration under a Changing Schema in Ampersand

### 2023-10-17 2022Migration/articleMigrationFACS.tex
Remove the example migration script.

\section{Generating a Migration Script}
   A migration script takes care of two things: copying data and ensuring that the migration system is a proper information system.
   Sections~\ref{sec:loading} and~\ref{sec:violations on invariants} illustrate how this works by example, using the Ampersand syntax.
   Then, section~\ref{General migration script} defines what this means in general terms.

\subsection{Loading Data}\label{sec:loading}
   Let us study how to migrate data from $\dataset$ to $\dataset'$.
   It would be tempting to state that a relation from the example system, say {\tt isPartOf}, could be copied by the following code fragment%
\footnote{The prefix {\tt old.} refers to concepts, relations, and rules from the existing system (section~\ref{sct:Example existing IS}).
The prefix {\tt new.} refers to items from the desired system.}:
\begin{verbatim}
   ENFORCE new.isPartOf >: old.isPartOf
\end{verbatim}
   The relation {\tt old.isPartOf} resides in the existing system and will not change after the MoT.

   This means, however, that anything in {\tt old.isPartOf} cannot be deleted from {\tt new.isPartOf}.
   This is undesirable because it prevents us from changing data in the migration system.
   The desired behavior is that all pairs in {\tt old.isPartOf} are copied to {\tt new.isPartOf} when the system is first initialized, but that {\tt new.isPartOf} can be changed freely after that.
   An extra relation (\ref{extraRelation}) achieves this behavior by remembering which pairs have been copied:
\begin{eqnarray}
   &&\verb#RELATION copied_isPartOf[Student*Course]#\label{extraRelation}\\
   &&\verb#ENFORCE new.isPartOf >: old.isPartOf - copied_isPartOf#\label{difference}\\
   &&\verb#ENFORCE copied_isPartOf >: new.isPartOf#\label{fill copies with new.isPartOf}
\end{eqnarray}

   Let us look what happens at runtime.
   Initially, when {\tt copied\_isPartOf} is still empty, the difference between {\tt old.isPartOf} and {\tt copied\_isPartOf} equals {\tt old.isPartOf}.
   So, {\tt new.isPartOf} will contain the pairs of {\tt old.isPartOf} by rule~\ref*{difference}.
   Then, the engine inserts the same pairs into {\tt copied\_isPartOf} because of rule~\ref*{fill copies with new.isPartOf}.
   After this, triples with the relation {\tt new.isPartOf} can be inserted and deleted freely.
   Once all triples have been copied, {\tt copied\_isPartOf} will contain all pairs of {\tt old.isPartOf} and rule~\ref*{difference} will no longer insert any triples in {\tt new.isPartOf}.
   So, this rule can be removed after the copying has taken place.
   Consequently, declarations~\ref{extraRelation} thru~\ref{fill copies with new.isPartOf} can be removed from the migration script at the MoC.

\subsection{Dealing with violations on invariants}\label{sec:violations on invariants}
   Suppose the script for the desired system adds an invariant that is not present in the existing system:

\begin{verbatim}
   RELATION isPartOf[Module*Course] [UNI]
\end{verbatim}

   The \verb=UNI= annotation requires that each \verb=Module= is part of at most one \verb=Course=.
   The violations are given by:
\begin{eqnarray}
   \viol{\tt UNI}{\dataset} = \{\pair{x}{y} \mid x \neq y \wedge \exists z.~ \pair{z}{x}, \pair{z}{y}\in\pop{\tt isPartOf}{\dataset} \}
\end{eqnarray}
   In Ampersand, the term that gives these violations is: \verb=isPartOf~;isPartOf - I=.

   While the \verb=UNI= requirement holds for the population in the original script, that population may have changed.
   Therefore, we must assume that we cannot expect that \verb=new.isPartOf= satisfies the \verb=UNI= requirement in the desired system.
   The migration system allows for this by simply not requiring that they hold at the MoT.
   A challenge lies in ensuring that they do hold at the MoC, and that users that help the migration forward can achieve a state in which all requirements hold.

   Two extra relations are used: one to track the violations that occur in the original data (\ref{violRelation}), and one to track the violations that have been fixed (\ref{fixedRelation}) which prevents them from recurring.
\begin{align}
   &\text{\tt RELATION viols\_UNI[Course*Course]}\label{violRelation}\\
   &\text{\tt RELATION fixed\_UNI[Course*Course]}\label{fixedRelation}\\
   &\text{\tt ENFORCE viols\_UNI >: old.isPartOf\textasciitilde;old.isPartOf - I}\label{fill viols}\\
   &\text{\tt ENFORCE fixed\_UNI >: viols\_UNI - (new.isPartOf\textasciitilde;new.isPartOf - I)}\label{fill fixed}\\
   &\begin{aligned}
   \text{\tt RULE isPartOf\_UNI: new.isPartOf\textasciitilde;new.isPartOf}\\
   \quad\text{\tt |- I \textbackslash/ (viols\_UNI - fixed\_UNI)}
   \end{aligned}\label{rule uni}
\end{align}

   We again look at what happens at runtime:
   Enforcement rule~\ref{fill viols} adds all violations present in the old system to {\tt viols\_UNI}.
   Until the MoT, rule~\ref{fill fixed} does nothing, since {\tt viols\_UNI} equals $\viol{\tt UNI}{\dataset}$.
   At that point, rule~\ref{rule uni} is satisfied, since the right-hand side of equation~\ref{rule uni} is empty.
   As violations are fixed, rule~\ref{fill fixed} records this by adding to {\tt fixed\_UNI}.
   This way, previously fixed errors cannot recur.
   Once {\tt viols\_UNI} is equal to {\tt fixed\_UNI}, the \verb=UNI= requirement holds and the system is ready for the MoC in regard to this rule.

   We also wish to alert users with the role `migration helper' to violations that need to be fixed.
   For the original \verb=UNI= requirement, this means specifying:

\begin{align}
   \begin{aligned}
   \text{\tt RULE isPartOf\_UNI\_warn: new.isPartOf\textasciitilde;new.isPartOf |- I}\\
   \quad\text{\tt ROLE MigrationHelper}
   \end{aligned}\label{warnings}
\end{align}

   Despite the syntax, Line~{\ref{warnings}} is not a rule of the generated system in the formal sense:
   By assigning a role, the responsibility of maintaining the rule no longer lies with the system.
   Instead, users with the migration helper role can see the violations and act on it as they see fit, so the only effect of Line~{\ref{warnings}} is that such users will be shown its violations.

\subsection{General migration script}\label{General migration script}
   This section presents a general approach for generating a migration script.
   Let $\overleftarrow{\infsys}$ be the existing system, and let $\overrightarrow{\infsys}$ be the desired system.
   Consequently, we define $\pair{\overleftarrow{\dataset}}{\overleftarrow{\schema}}=\overleftarrow{\infsys}$ and $\pair{\overrightarrow{\dataset}}{\overrightarrow{\schema}}=\overrightarrow{\infsys}$ as the data set and schema of the existing and desired system, respectively.
   We also define $\quadruple{\overleftarrow{\concepts}}{\overleftarrow{\rels}}{\overleftarrow{\rules}}{\overleftarrow{\transactions}}=\overleftarrow{\schema}$
   and $\quadruple{\overrightarrow{\concepts}}{\overrightarrow{\rels}}{\overrightarrow{\rules}}{\overrightarrow{\transactions}}=\overrightarrow{\schema}$.

@@

   Let $\pair{\dataset}{\schema}$ be the existing system, and let ${\schema'}$ be the schema of the desired system.
   Then we define the migration systems $\migrsys$ and the events as follows:
\begin{enumerate}
\item We take a disjoint union of the relations, defining $\overleftarrow~$ and $\overrightarrow~$ as (overloaded) notation to generate new names, keeping the data set of the old system and the rules of the new system:
   \begin{align}
   \overleftarrow{\declare{n}{x}{y}}&=\declare{(\id{old}.n)}{x}{y}&\overleftarrow{X} &= \{\overleftarrow{x}\mid x\in X\}\\
   \overrightarrow{\declare{n}{x}{y}}&=\declare{(\id{new}.n)}{x}{y}& \overrightarrow{X} &= \{\overrightarrow{x}\mid x\in X\}\\
   \triples_\migrsys &= \{\triple{a}{\overleftarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\dataset\}&
   \instance_\migrsys &= \instance_\dataset\\
   \rels_\theenumi &= \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'}
   %&\rules_\theenumi &= \rules_\schema
   \end{align}
\item We create transactions to copy equal relations, corresponding to equation~\ref{extraRelation} through~\ref{fill copies with new.isPartOf} in Section~\ref{sec:loading}:
   \begin{align}
   \widetilde{\declare{n}{x}{y}} &= \declare{(\id{copied}.n)}{x}{y}\\
   \rels_\theenumi &= \{\widetilde{r}\mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
   \transactions_\theenumi &= \{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid \widetilde{r} \in \rels_2\} \cup \{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}} \mid \widetilde{r} \in \rels_2\}
   \end{align}
\item For each rule, we generate two helper relations, {\tt viols} and {\tt fixed}. Note that Rule~\ref{warnings} only influences the user interface and therefore is not modeled here. This corresponds to equation~\ref{violRelation} through~\ref{rule uni}.
   \begin{align}
   \overleftarrow{\dataset'} &= \pair{\{\triple{x}{{r}}{y}\mid \triple{x}{\overleftarrow{r}}{y} \in \triples_{\dataset'}\}}{\instance_{\dataset'}}\\
   \overrightarrow{\dataset'} &= \pair{\{\triple{a}{{r}}{b} \mid \triple{a}{\overrightarrow{r}}{b}\in\triples_\dataset\}}{\instance_{\dataset'}} \\
   \rels_\theenumi &= \{{\tt viols}_u\mid u \in \rules_{\schema'}\} \cup \{{\tt fixed}_u\mid u \in \rules_{\schema'}\}\\
   \rules_\migrsys &= \{\lambda {\dataset'}.~ \viol{u}{\overrightarrow{\dataset'}} - \pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'} \mid u\in\rules_{\schema'}\}\label{eqn:migrsysrules}\\
   \transactions_{\theenumi} &=\{{\tt viols}_u \mapsfrom \lambda \dataset'.~ \viol{u}{\overleftarrow{\dataset'}}\} \cup
    \{{\tt fixed}_u \mapsfrom \lambda \dataset'.~ {\tt viols}_u - \viol{u}{\overrightarrow{\dataset'}}\}\label{eqn:enforceForRules}
   \end{align}
\item Combining the above into a single migration system:
   \begin{align}
   \concepts_\migrsys &= \concepts_\dataset \cup  \concepts_{\dataset'}\\
   \rels_\migrsys &= \rels_1\cup\rels_2\cup \rels_3\\
   %\rules_\migrsys &= \rules_1\cup \rules_3\\
   \transactions_\migrsys &= \transactions_1\cup\transactions_2\cup\transactions_3
   \end{align}
\end{enumerate}

>>>>

\section{Generating a Migration Script}
   A migration script takes care of two things: copying data and ensuring that the migration system is a proper information system.
   This section presents a general approach for generating a migration script.
   Let $\overleftarrow{\infsys}$ be the existing system, and let $\overrightarrow{\infsys}$ be the desired system.
   Consequently, we define $\pair{\overleftarrow{\dataset}}{\overleftarrow{\schema}}=\overleftarrow{\infsys}$ and $\pair{\overrightarrow{\dataset}}{\overrightarrow{\schema}}=\overrightarrow{\infsys}$ as the data set and schema of the existing and desired system, respectively.
   We also define $\quadruple{\overleftarrow{\concepts}}{\overleftarrow{\rels}}{\overleftarrow{\rules}}{\overleftarrow{\transactions}}=\overleftarrow{\schema}$
   and $\quadruple{\overrightarrow{\concepts}}{\overrightarrow{\rels}}{\overrightarrow{\rules}}{\overrightarrow{\transactions}}=\overrightarrow{\schema}$.

@@

   Let $\pair{\dataset}{\schema}$ be the existing system, and let ${\schema'}$ be the schema of the desired system.
   Then we define the migration systems $\migrsys$ and the events as follows:
\begin{enumerate}
\item We take a disjoint union of the relations, defining $\overleftarrow~$ and $\overrightarrow~$ as (overloaded) notation to generate new names, keeping the data set of the old system and the rules of the new system:
   \begin{align}
   \overleftarrow{\declare{n}{x}{y}}&=\declare{(\id{old}.n)}{x}{y}&\overleftarrow{X} &= \{\overleftarrow{x}\mid x\in X\}\\
   \overrightarrow{\declare{n}{x}{y}}&=\declare{(\id{new}.n)}{x}{y}& \overrightarrow{X} &= \{\overrightarrow{x}\mid x\in X\}\\
   \triples_\migrsys &= \{\triple{a}{\overleftarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\dataset\}&
   \instance_\migrsys &= \instance_\dataset\\
   \rels_\theenumi &= \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'}
   %&\rules_\theenumi &= \rules_\schema
   \end{align}
\item We create transactions to copy equal relations:
   \begin{align}
   \widetilde{\declare{n}{x}{y}} &= \declare{(\id{copied}.n)}{x}{y}\\
   \rels_\theenumi &= \{\widetilde{r}\mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
   \transactions_\theenumi &= \{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid \widetilde{r} \in \rels_2\} \cup \{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}} \mid \widetilde{r} \in \rels_2\}
   \end{align}
\item For each rule, we generate two helper relations, {\tt viols} and {\tt fixed}.
   \begin{align}
   \overleftarrow{\dataset'} &= \pair{\{\triple{x}{{r}}{y}\mid \triple{x}{\overleftarrow{r}}{y} \in \triples_{\dataset'}\}}{\instance_{\dataset'}}\\
   \overrightarrow{\dataset'} &= \pair{\{\triple{a}{{r}}{b} \mid \triple{a}{\overrightarrow{r}}{b}\in\triples_\dataset\}}{\instance_{\dataset'}} \\
   \rels_\theenumi &= \{{\tt viols}_u\mid u \in \rules_{\schema'}\} \cup \{{\tt fixed}_u\mid u \in \rules_{\schema'}\}\\
   \rules_\migrsys &= \{\lambda {\dataset'}.~ \viol{u}{\overrightarrow{\dataset'}} - \pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'} \mid u\in\rules_{\schema'}\}\label{eqn:migrsysrules}\\
   \transactions_{\theenumi} &=\{{\tt viols}_u \mapsfrom \lambda \dataset'.~ \viol{u}{\overleftarrow{\dataset'}}\} \cup
    \{{\tt fixed}_u \mapsfrom \lambda \dataset'.~ {\tt viols}_u - \viol{u}{\overrightarrow{\dataset'}}\}\label{eqn:enforceForRules}
   \end{align}
\item Combining the above into a single migration system:
   \begin{align}
   \concepts_\migrsys &= \concepts_\dataset \cup  \concepts_{\dataset'}\\
   \rels_\migrsys &= \rels_1\cup\rels_2\cup \rels_3\\
   %\rules_\migrsys &= \rules_1\cup \rules_3\\
   \transactions_\migrsys &= \transactions_1\cup\transactions_2\cup\transactions_3
   \end{align}
\end{enumerate}
